Ensemble Transport Adaptive Importance Sampling

### 2019-01-09 response_JUQ_2.tex

\usepackage[colorlinks]{hyperref}

\newcommand{\comment}[2]{\vspace{0.6cm}{\bf Comment:} {\it #1.}

\vspace{0.3cm}{\bf Answer:} #2}

\begin{document}
\title{Response to Referee Comments: Parallel Adaptive Importance Sampling}
\maketitle
First of all, we would like to thank the referees and the associate editor for
their time in reading and commenting on our manuscript. We thank all of you and Professor Higdon for your patience in waiting for these delayed corrections. In this document
we will address each of the points raised, and describe any relevant
changes that have been made to the paper. We have made several substantial changes, including the renaming of the algorithm and paper from ``parallel adaptive importance sampling'' (PAIS) to ``ensemble transport adaptive importance sampling'' (ETAIS). We have also replaced two of the numerical examples. In particular we have incorporated a new bimodal example with differing scales in the two modes, and for which we have tested the ETAIS with three different resampling algorithms. This in particular we feel makes the paper stronger, and we hope that the referees and editors will now agree that the paper is ready for publication.

@@

\section*{Referee 1 comments}

\comment{This is supposed to be a parallel algorithm and yet it's implemention is serial. Furthermore, the authors don't seem to be entirely sure that parallelization will work since they comment that it's effectiveness will depend on various factors. The title seems somehow misleading to me. Maybe if the "parallel" part of the title is removed and the discussion about parallelization is treated as a possibility instead of a critical part of the algorithm, then the document will be less confusing.}{Despite the fact that we felt that we had demonstrated the feasibility of the parallelisation of this method, we agree with the referee that we have not demonstrated this in the numerics. Furthermore, this algorithm delivers a speed-up even when run in serial. As such, we are happy to change the name of the algorithm to ``Ensemble Transport Adaptive Importance Sampling'' (ETAIS). This required some rewriting throughout the paper. In particular, we have toned down the emphasis on parallelisation, and put more emphasis on the fact that this algorithm also delivers speed-up for serial implementations. See the second paragraph of the discussion.}

\comment{Also, the comparison between serial PAIS and the naive parallel MCMC is strange if there is no parallelization.}{Most of the numerics compares our methodology to Metropolis-Hastings, based on the number of likelihood evaluations required, rather than runtime, with the argument being based on the idea that this approach is most suitable for targets with expensive likelihood evaluations. Since the only difference between implementations in parallel or serial of our method will be in CPU runtime (due to extra overheads due to communications between processors), it is still valid for us to compare to a naively parallelised Metropolis-Hastings approach, since our choice of serial/parallel implementation makes no difference to these results. For the final example where we consider runtime, a sentence has been added to state that our results will be slightly overoptimistic based on the lack of communication overheads.}

\comment{It will be interesting that the authors show the PAIS using AMS with the examples. They show the difference between the AMS and the ETPF but not within the PAIS.}{We assume that the referee meant the AMR? As the AE asked for a new example to replace the second example, we took this as an opportunity to compare ETAIS with the ETPF, AMR and bootstrap resamplers, which is now presented in the new first example. The numerics demonstrate that superior resamplers lead to more stable and accurate importance sampling, with ETPF algorithm narrowly outperforming the AMR algorithm, with the bootstrap version following up in the rear. This helps support our claim that state-of-the-art resampling methods can lead to more efficient AIS algorithms.}

@@

\comment{3) It might be trivially to do so, but the authors only briefly mention a prove for the convergence of their algorithm. In fact, is it ok to use a dependent proposal in importance sampling? ... no regularity conditions are needed? Why Alg. 3 works? resampling "introduces error" ... is this relevant? does the error cascades down in each iteration? In the examples the PAIS seem to work ok, but, do we need any regularity condition? Please make your convergence claims more explicit and take more time to explain them. In particular, do we have the correct invariant distribution?}{This algorithm belongs to a family of methods for which convergence has been proven in the literature, and as such, we have referenced this proof, and as such we did not feel it appropriate to provide further details. One detail was missing from our statement, which is that the posterior distribution must be absolutely continuous with respect to the proposal distribution, which we have now added.

Regarding the error mentioned in the comment, I believe the AE is referring to a comment regarding the fact that the importance samples are stored for Monte Carlo estimates rather than the resampled samples, since the resampling process is not exact. Any introduction of errors in the resampling process simply leads to a proposal distribution which is less representative of the target distribution, and as such the resulting importance sample will have larger variance in the weights. The method will still converge to the correct distribution, but larger variance in the weights leads to a less efficient method. We have added a paragraph to clarify this.}

\comment{4) There are too many examples. The univariate gaussian is not very informative and the univariate bimodal, is not either since both modes are of the same scale. I would rather see these two examples turned into one single one with a 2D mixture of two separated gaussians with contrasting scales (variance covariance matrices).}{We agree that the numerical experiments section is very long. However the examples presented do each demonstrate a different aspect of the advantages of our approach. We have taken this comment on board, however, and have drastically reduced the length of example 1, in which we demonstrate that tuning the proposal variances using the effective sample size is approximately equivalent to using the $L^2$ errors in the target, which is only available to us as we have picked a trivial target. We have replaced the second bimodal example, with a bimodal example as requested by the AE with differing scales in the covariances of the two modes. We have also taken this opportunity to demonstrate the differences in results when using the different resamplers discussed in the paper. We have also included a comparison with a standard bootstrap resampler. The new numerics demonstrate the improvement in stability and convergence of the method with the use of more advanced resampling methods. The mixture model example demonstrates the algorithm working in a higher dimensional multimodal setting on a popular practical problem. The final new example (see later comments) demonstrates the speed-ups achieved using our approach for a challenging problem with an expensive likelihood, requiring the approximation of a differential equation. We believe that each example serves a purpose, but we have also reduced the size of this section by several pages.}

@@

\comment{5) In the examples the authors include very ad hoc strategies like the 'scout' chain and specially the very ad hoc strategy described in section 7.4.2 (!). The authors, in my opinion, do not really discuss the implementation of the PAIS algorithm but a case-by-case ad hoc version of it. This is clearly unacceptable.}{The scout chains were only used in the old bimodal example, and have since been cut. This approach has already been used in the pMC literature to stabilise the importance sampling by fattening the tails of the proposal distribution, but we have no problem with removing it from this paper.

  As for the proposal kernels used in 7.4.2, we disagree that this approach is ``ad-hoc''. These are chosen simply because the proposal distribution must only be supported by [0,1], and the distribution chosen give us a way to make proposals which are sure to be in the this interval but with tuneable mean and variance. This relates to the previous answers we have given to comments made by the referee above. We have gone into more details regarding this and explaining the choices.}

@@

\comment{6) Please clarify the first phrase of section 7.4.2, which, as far as I see, is simply twice wrong. There is a (very simple!) analytic version of the posterior AND with the right proposal MH can sample from it (!).}{We have removed this phrase from the paper.}
% {We have amended the statement about the analytic form of the posterior. We are not aware of any M-H methods which can reliably sample from multimodal densities with well-separated modes which are not known a priori. HMC trajectories can be hand-tuned to make transitions between modes, but only with information regarding the separation. Ensemble-based methods appear to us to be a good candidate for reliable sampling from such distributions without such a hand-tuning. However, we agree to remove the claim.}

\comment{7) Although the chaotic Lorenz 93 ODE is used, it is used in such a short time scale that the posterior is quite well behaved (unimodal and fairly gaussian and only some correlation is seen between to parameters). Since the priors are normal, this regime suggest a linear part of the FM. The authors should attempt something more challenging, by running the system into a highly nonlinear regime. Otherwise, why consider this example?}{The reason for considering the example was that the AE had previously asked us to produce an example in the regime where we suggest that we get the biggest win using our proposed method, i.e. those where the likelihood evaluations have significant cost due to requirement of, for example, the solution of a differential equation. In this example, we have demonstrated a significant speed-up over the equivalent M-H algorithm. We do understand however the AE's desire for a more challenging inverse problem in itself, and so we have created a new example which has a prior distribution with larger variance. The push-forward of this prior distribution is multimodal on the timescale of the observations (see the figure below), showing that trajectories do pass close to a hyperbolic point in the assimilation window. This also results in a longer thinner posterior, which as can be seen from the numerics, RWMH really struggles to sample from efficiently. The ETAIS implementation handles this problem with ease, for a big speed-up. The posterior is challenging because it is concentrated very close to a lower dimensional manifold. Overall, the example demonstrates the speed-up of our approach for a problem where the likelihood is expensive.

@@

\begin{figure}[htp]
\includegraphics[width=0.3\textwidth]{figures/xhist}
\includegraphics[width=0.3\textwidth]{figures/yhist}
\includegraphics[width=0.3\textwidth]{figures/zhist}
\caption{Histograms from 1e7 Monte Carlo runs of the push forward of the new prior for the Lorenz '63 problem}
\end{figure}

The more dimensions and posterior complexity that exists in a problem, the bigger the ensemble will need to be to have a stable ETAIS implementation. After a certain point, due to the extra overheads with a large ensemble due to for example the resampling, ETAIS will no longer be competitive. This said, in our most recent work, arbitrarily complex posteriors can be efficiently sampled using ETAIS if used in conjunction with a transport map to simplify the target. This is the subject of our next paper which we will be submitting imminently. We have added a paragraph to this end to the discussion.}

### 2019-01-09 ETAIS.tex
Draft complete
Done I hope!!!!!!!!!!!

{\red Figure \ref{fig:Lorenz_convergence} shows convergence plots for
  this problem using RWMH and ETAIS. The burn-in of the ETAIS is
  remarkably better than MH, as the ensemble quickly covers the
  manifold on which the majority of the density lies close to. The
  plot shows convergence with respect to number of samples, but the
  two approaches have differing costs, due to the calculation of the
  denominator in the weights in the ETAIS, and the resampling step
  (AMR). Assuming that both error curves decay like $N^{-1/2}$ from
  $N=10^6$, then for a given error tolerance, the number $N_{\rm MH}$
  of MH samples required is given by
\[N_{\rm MH} = N_{\rm ETAIS} \times 3.838 \times 10^4,\]
where $N_{\rm ETAIS}$ is the number of ETAIS samples required for the
same error tolerance. The cost per sample for the MH and ETAIS methods
for this problem are $1.64 \times 10^{-5}$ and $1.46 \times 10^{-3}$
respectively. Putting this together, we arrive at a speed-up factor of
over 430. This demonstrates the benefits of this approach for
problems with expensive likelihoods and challenging posteriors such as
this one.
}

@@

However, tuning the variances and
covariances of the mixture proposal components
globally is likely to be of limited use for multimodal problems where
the modes have very different covariances, or where there are curved
ridges in the density. There are a variety of possible solutions to
this problem that would be worthy of future consideration, not least
the potential to use transport maps\cite{el2012bayesian,parno2014transport} to map Gaussian mixtures to
highly complex non-Gaussian approximations of the posterior. Such a
map could encode local covariance information, and lead to accelerated
and stable ETAIS sampling. This is an interesting area for future study.

>>>>

{\red Figure \ref{fig:Lorenz_convergence} shows convergence plots for
  this problem using RWMH and ETAIS. The burn-in of the ETAIS is
  remarkably better than MH, as the ensemble quickly covers the
  manifold on which the majority of the density lies close to. The
  plot shows convergence with respect to number of samples, but the
  two approaches have differing costs, due to the calculation of the
  denominator in the weights in the ETAIS, and the resampling step
  (AMR). Assuming that both error curves decay like $N^{-1/2}$ from
  $N=10^6$, then for a given error tolerance, the number $N_{\rm MH}$
  of MH samples required is given by
\[N_{\rm MH} = N_{\rm ETAIS} \times 3.838 \times 10^4,\]
where $N_{\rm ETAIS}$ is the number of ETAIS samples required for the
same error tolerance. The cost per sample for the MH and ETAIS methods
for this problem are $1.64 \times 10^{-5}$ and $1.46 \times 10^{-3}$
respectively. Here the ETAIS results were produced with a serial
implementation, so this cost may be slightly overoptimistic, since the
runtimes do not include extra overheads of communicating the ensemble
states between processors. However, putting this together, we arrive at a speed-up factor of
over 430, which will undoubtedly overshadow any such underestimate of
the cost-per-sample of the ETAIS approach. This demonstrates the benefits of this approach for
problems with expensive likelihoods and challenging posteriors such as
this one. 
}

@@

However, tuning the variances and
covariances of the mixture proposal components
globally is likely to be of limited use for multimodal problems where
the modes have very different covariances, or where there are curved
ridges in the density. {\red With increased dimensionality and/or
  complexity of the target the distribution, we also require increases
  in the size of the ensemble if we wish to have a stable ETAIS
  implementation. After a certain point, the extra overheads
  associated with a very large ensemble will outweigh the advantages
  of this approach. However, there} are a variety of possible solutions to
this problem that would be worthy of future consideration, not least
the potential to use transport maps\cite{el2012bayesian,parno2014transport} to map Gaussian mixtures to
highly complex non-Gaussian approximations of the posterior. Such a
map could encode local covariance information, and lead to accelerated
and stable ETAIS sampling {\red with much smaller ensemble sizes}.
